---
title: "02 - Data Manipulation"
author: "David Vichansky [6819516]"
date: "12-11-2020"
mainfont: Garamond
fontsize: 12pt
urlcolor: blue
output: 
  pdf_document:
    latex_engine: xelatex
---

Here is an example file you can write.

First, load the packages:

```{r load_packages}
library(ISLR)
library(tidyverse)
library(haven)
library(readxl)
library(tinytex)
```

1. Objects and classes

```{r objects}
object_1 <- 1:5
object_2 <- 1L:5L
object_3 <- "-123.456"
object_4 <- as.numeric(object_2)
object_5 <- letters[object_1]
object_6 <- as.factor(rep(object_5, 2))
object_7 <- c(1, 2, 3, "4", "5", "6")
```

```{r classes}
df <- data.frame(object_1, object_2) 
sapply(df, class)
# creating dataframe for all objects does not work due to differing row lengths
class(object_1)
class(object_2)
class(object_3)
class(object_4)
class(object_5)
class(object_6)
class(object_7)
# Here is a comment about the class of object_1
```

2. Convert to numeric

```{r convert to numeric}
object_7 <- as.numeric(object_7)
class(object_7)
```

3. Create list titled 'objects'

```{r create list}
objects <- list(object_1,object_2,object_3,object_4,object_5,object_6,object_7)
objects
```

4. Create data-frame

```{r list}
dataframe <- data.frame(object_1, object_2, object_5)
dataframe
```

5. Determine size of data-frame

```{r size}
ncol(dataframe)

nrow(dataframe)
```

6. Read csv file

```{r csv}
apps <- read.csv("data/googleplaystore.csv", header = TRUE)
#too large to print in pdf
#apps
```

7. Did any column get a variable type you did not expect?
```{r column varaibles}
lapply(apps, class)
```
We suppose the variable type "factor" is unexpected in the sense we haven not worked with this yet in this example, "numeric" and "integer" we know the functionalities of those two data types.

8. First row of data-frame 'apps'.
```{r headings}
head(apps)
```

9. Repeating steps '5', '6' and '7' for another data set.
```{r repeat}
# load data set
student <- read.csv("data/students.csv", header = TRUE)
#too large to print in pdf
#student

# check dimension of dataframe
ncol(student)

nrow(student)

# check 'class' of variables
lapply(student, class)

# 'tail' and 'View' of data set
tail(student)

View(student) # opens data set in seperate window
```

10. Create summary of 'student' data set.
```{r summary}
summary(student)
```

11. Filter students with a grade lower than 5.5.
```{r filter}
filter(student, grade < 5.5)
```

12. Filter for students with a grade higher than 8 from programme A.
```{r filter2}
filter(student, grade > 8.0, programme == "A")
```

13. Sort students from programme A are on top of the data frame and within the programmes the highest grades come first.
```{r sort}
arrange(student, programme, -grade)
```

14. Show only the student_number and programme columns from the students dataset.
```{r show}
select(student, ï..student_number, programme)
```

15. Change the codes in the programme column of the students dataset to their names.
```{r recode}
students_recoded <- student %>% mutate(programme=recode(programme, 
                         `A`="Science",
                         `B`="Social Science"))
#too large to print
# students_recoded
```

16. Create a data processing pipeline that (a) loads the apps dataset, (b) parses the number of installs as ‘Downloads’ variable using mutate and parse_number(), (c) shows only apps with more than 500 000 000 downloads, (d) orders them by rating (best on top), and (e) shows only the relevant columns (you can choose which are relevant, but select at least the Rating and Category variables). Save the result under the name popular_apps.
```{r appsdata}
popular_apps <- read.csv("data/googleplaystore.csv") %>%
  filter(Installs == "500,000,000+") %>%
  mutate(Downloads=parse_number("500,000,000+")) %>%
  arrange(Rating) %>% 
  select(Rating, Category, Content.Rating, Genres)

#too large to print
# popular_apps
```

17. Show the median, minimum, and maximum for the popular apps dataset.
```{r summar}
popular_apps %>% 
  summarise(
    mean = mean(Rating), 
    variance = var(Rating), 
    min = min(Rating), 
    max = max(Rating)
  )
```

18. Add the median absolute deviation to the summaries.
```{r mad}
mad <- function(x) {
  median(abs(x - median(x)))
}

popular_apps %>% summarise(mad = mad(Rating))
```

19. Create a grouped summary of the ratings per category.
```{r summariseratings}
popular_apps %>% 
  group_by(Category) %>% 
  summarise(
    mean = mean(Rating), 
    variance = var(Rating), 
    min = min(Rating), 
    max = max(Rating)
  )

```

20. Create a summary based on the Google play store apps dataset. We create a pivot table.
```{r meanratings}
#popular123 <- popular_apps %>% 
#  group_by(Category, Content.Rating) %>% 
#  summarise(
#    mean = mean(Rating)
#  )
#
#pivot_wider(popular, id_cols = Category, names_from = 'Content.Rating', values_from = mean) #%>%


# create pivot table to show the mean rating by app 'category' to then compare across age group ('content.rating')
popular_apps %>% 
  group_by(Category, Content.Rating) %>% 
  summarise(
    mean = mean(Rating)
  ) %>%
  pivot_wider(id_cols = Category, names_from = 'Content.Rating', values_from = mean)
```
