---
title: "Assignment Prediction Model"
author: "Manousos Emmmanouil Theodosiou [6686311], Group: GG Force"
date: "11-01-2021"
output:
  html_document:
    df_print: paged
  pdf_document:
    latex_engine: pdflatex
fontsize: 12pt
urlcolor: blue
mainfont: Arial
---

```{r}
library(MASS)
library(class)
library(ISLR)
library(tidyverse)
library(rpart)
library(rpart.plot)
library(randomForest)
```

Load the dataset "accidents_2012_to_2014.csv"

```{r}
accidents <-
  read_csv("accidents_2012_to_2014.csv") 
#View(accidents)

#distinct(accidents, Road_Type)
#distinct(accidents, Weather_Conditions)
#distinct(accidents, Light_Conditions)
#distinct(accidents, Road_Surface_Conditions)
#distinct(accidents, Special_Conditions_at_Site)
#distinct(accidents, Carriageway_Hazards)
```

By looking at the dataset it can be seen that in 2012-2014 the most accidents occur on Saturday whereas the least number of accidents take place on Monday. (1 is Monday, 2 is Tuesday, etc). 
```{r}
accidents %>% 
  group_by(Day_of_Week) %>% 
  summarize(total_accidents=n_distinct(Accident_Index)) %>%
    ggplot(aes(x=Day_of_Week, y=total_accidents)) +
    geom_bar(stat="identity", fill="light seagreen")+
    geom_text(aes(label=total_accidents), vjust=1.6, color="white", size=3.5)+
    theme_minimal()
```

We can also visualise the number of accidents by hour. We can see that most accidents take place around 5pm and 8am. This could make sense since people having 9 to 5 jobs are more likely to have an accident on their way to or from their jobs. 

```{r}
accidents %>% 
  mutate(time_slot = as.numeric(substr(Time,0,2))) %>% 
  group_by(time_slot) %>% 
  summarize(total_accidents=n_distinct(Accident_Index)) %>%
    ggplot(aes(x=time_slot, y=total_accidents)) +
    geom_bar(stat="identity", fill="light seagreen")+
    geom_text(aes(label=total_accidents), vjust=1.6, color="black", size=3)+
    scale_x_continuous(breaks = round(seq(0, 24, by = 2),0)) +
    ggtitle("Total Accidents by Hours") +
    xlab("Hours") + ylab("Total Accidents")+
    theme(plot.title = element_text(hjust = 0.5), panel.background = element_blank())
```
Here, we will analyse the probability of having an accident (1 for Fatal, 2 for Serious, 3 for Slight) in a given hour.

```{r}
accidents_time <- 
  accidents %>% 
  mutate(time_slot = as.numeric(substr(Time,0,2)))

accidents_time_severity <- table(accidents_time$time_slot, accidents_time$Accident_Severity)
prop.table(accidents_time_severity,1)
```

```{r}
View(accidents_time)
```


Now we will choose only the relevant predictors for the response variable Accident_Severity and make a dataframe.
```{r}
variables <- names(accidents_time) %in% c("Accident_Severity", 
                                 "Number_of_Vehicles", 
                                 "Number_of_Casualties",
                                 "Day_of_Week",
                                 "time_slot",
                                 "Road_Type",
                                 "Speed_limit",
                                 "Light_Conditions",
                                 "Weather_Conditions",
                                 "Road_Surface_Conditions"
                                 )

accidents_df <- data.frame(accidents_time[variables])
View(accidents_df)
```

Check the type of each column by running the sapply() function.

```{r}
sapply(accidents_df, class)
```

Make classification trees using the rpart() function.

Using all predictors:
```{r}
set.seed(1)
accidents_tree_mod <- rpart(Accident_Severity ~ ., data = accidents_df, 
                            control=rpart.control(minsplit=10, cp=0.001))
rpart.plot(accidents_tree_mod)
```

Using Accident_Severity ~ Road_Type + Light_Conditions + Weather_Conditions + Road_Surface_Conditions:

```{r}
set.seed(1)
accidents_tree_mod_1 <- rpart(Accident_Severity ~ Road_Type + Light_Conditions
                              + Weather_Conditions + Road_Surface_Conditions,
                              data = accidents_df, 
                              control=rpart.control(minsplit=10, cp=0.001))
rpart.plot(accidents_tree_mod_1)
```


## Random Forest for classification (RUNS OUT OF MEMORY)

```{r}
#accidents_rf <- randomForest(Accident_Severity ~ ., data = accidents_df, importance = TRUE)
#accidents_rf
```

Importance
```{r}
#importance(accidents_rf)
```

Plot importance
```{r}
#varImpPlot(accidents_rf)
```

